A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-12-20 podc_paper/algorithm.tex
working on description of block fields

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
Instead, the block stores an implicit representation of the sequences.

>>>>

Each node has an infinite array called \var{blocks}.
To simplify the code, each node has an empty block in \var{blocks[0]}.

If a block contained an explicit representation of its sequences of enqueues and dequeues,
it would take $\Omega(p)$ time to construct a block (task \ref{construct}), which would be much too slow.
Instead, the block stores an implicit representation of the sequences; see Figure \ref{object-fields}.

\here{This parag may need to be rewritten if we use sum\sub{enq} field instead of sum\sub{enq-left} and sum\sub{enq-right}.  Also, maybe it is too detailed at this point; perhaps instead just say that for task \ref{findenqueue} we want to be able to find the position of the $r$th enqueue in a block within a block in its child using binary search.}
Each block in an internal node $v$ stores the indices of the last direct subblock in $v$'s left and right child in the fields \eleft\ and \eright.  This allows us to find the direct subblocks of any block easily.
Now, suppose we want to find the $r$th enqueue in the linearization ordering for task \ref{findenqueue}.
Let $B_1, B_2, \ldots, B_k$ be the blocks in the root.
First, we need prefix sums of the number of enqueues in $E(B_1)\cdot E(B_2)\cdots E(B_i)$
so that we can do a binary search for the block $B_e$ that contains the $r$th enqueue.
This prefix sum also allows us to know the rank $r'$ within $E(B_e)$ of the $r$th enqueue.
Once we have $r'$, we need the number of enqueues that $B_e$ received from its left child
to determine whether the enqueue came from the left or right child of the root.
Suppose the enqueue came from the right child $v_r$.
Then, we know that the index of the block in $v_r$ that contains the enqueue
is between $B_{e-1}.\eright + 1$ and $B_e$.\eright.
We can again do a binary search within this range.
For this, we can again use the prefix sums of the number of enqueues in any prefix of the array $v_r.blocks$.
We can then continue in this way down the tree until reaching a leaf where the enqueue is stored explicitly.
We shall show that the binary search in the root can be done in $O(\log p + \log q)$ steps,
and the binary search within each other node along the path to a leaf can be done in $O(\log p)$ steps,
for a total of $O(\log^2 p + \log q)$ steps to complete task \ref{findenqueue}.
All of the information for this search process can be derived from the 
\var{sum\sub{enq-left}} and \var{sum\sub{enq-right}} fields.

To facilitate \ref{findinroot}, each block has a field \var{super} that contains
the (approximate) index of its superblock in the parent node's blocks array.
We will ensure that this field's value differs from the true index of the superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two blocks in the parent node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation all the way up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ time.

\begin{figure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Block}}

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
\begin{itemize}
    \item \tt{int} \eleft, \eright
  \textsf{\com\ indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
\begin{itemize}
  \item \tt{\sl{int} \size}%
  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
\end{itemize}

\end{algorithmic}
\caption{Objects used in Tournament Tree Data Structure \label{object-fields}}
\end{figure}

### 2022-12-20 podc_paper/podc-submission.tex
working on description of block fields

@@

\begin{algorithm}
\caption{Objects used in Tournament Tree Data Structure}
\begin{algorithmic}[1]
\setcounter{ALG@line}{1}

\Statex $\diamondsuit$ \tt{\sl{Shared}}
\begin{itemize}
\item \tt{\sl{Node} root} \sf{\com\ the root of a binary tree of \tt{Node}s with one leaf for each process}
\end{itemize}

\Statex $\diamondsuit$ \tt{\sl{Local}}
\begin{itemize}
\item \tt{\sl{Node} leaf} \sf{\com\ process's leaf in the tree}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{Node} left, right, parent} \textsf{\com\ children and parent pointers initialized  when creating the tree}
\item \tt{\sl{Block[0..$\infty$]} blocks} \textsf{\com\ initially \tt{blocks[0]} contains an empty block with all integer fields equal to 0}
\item \tt{\sl{int} \head} \textsf{\com\ position to attempt appending next \tt{block} to \tt{blocks}, initially 1}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{Block}}

\begin{itemize}
  \item \tt{\sl{int} super}
  \textsf{\com\ approximate index of the block's superblock in \var{parent.blocks}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{InternalBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in internal nodes}
\begin{itemize}
    \item \tt{\sl{int} end\sub{left}, end\sub{right}}
  \textsf{\com\ indices of the last subblock of the block in the left and right child}
  \item \tt{\sl{int} sum\sub{enq-left}}\textsf{\com\ number of enqueues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{deq-left}}\textsf{\com\ number of dequeues in \tt{left.blocks[1..end\sub{left}]}}
  \item \tt{\sl{int} sum\sub{enq-right}}\textsf{\com\ number of enqueues in \tt{right.blocks[1..end\sub{right}]}}
  \item \tt{\sl{int} sum\sub{deq-right}}\textsf{\com\ number of dequeues in \tt{right.blocks[1..end\sub{right}]}}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{LeafBlock} extends \sl{Block}} \sf{\com\ the following additional fields are used only for blocks in leaves}
\begin{itemize}
  \item \tt{\sl{Object} element}
  \textsf{\com\ if the operation stored in the block is \tt{enqueue(x)} then \tt{element=x}, otherwise \tt{element=null}.}

    \item \tt{\sl{int} sum\sub{enq}, sum\sub{deq}}
  \textsf{\com\ number of enqueue, dequeue operations in  leaf's \var{blocks} array up to this block (inclusive)}
\end{itemize}

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}} \sf{\com\ the following field is used only for blocks in the root}
\begin{itemize}
  \item \tt{\sl{int} \size}%
  \textsf{\com\ size of the queue after performing all operations up to the end of this block in linearization order}
\end{itemize}

@@

\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{new.end\sub{dir}= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].end\sub{dir}..dir.\head-1]}.}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \tt{block\sub{last}= dir.blocks[new.end\sub{dir}]}
\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
\EndIf

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].end\sub{left}+1,} \label{leftChildGet}
\Statex \hspace{10.7em}\tt{blocks[b].end\sub{left})} 
\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].end\sub{right}+1,} \label{rightChildGet}
\Statex \hspace{11.1em}\tt{blocks[b].end\sub{right})} 
\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
\EndIf
\EndFunction{GetEnqueue}

>>>>

\newcommand{\eleft}{\var{end\sub{left}}}
\newcommand{\eright}{\var{end\sub{right}}}
\newcommand{\edir}{\var{end\sub{dir}}}

@@

@@

\Function{Block}{CreateBlock}{\sl{int} i} \cmt{Creates and returns the block to be installed in \tt{blocks[i]}.}
\State \tt{block new= \Call{new}{\sl{InternalBlock}}} \label{initNewBlock}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{prev}= blocks[i-1].\edir} \label{prevLine}
\State \tt{new.\edir= dir.\head-1} \label{lastLine} \cmt{\nf{new} contains \tt{dir.blocks[blocks[i-1].\edir..dir.\head-1]}.}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \tt{block\sub{last}= dir.blocks[new.\edir]}
\State \tt{new.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}} \label{setSumEnqLeft}
\State \tt{new.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}} \label{setSumEnqRight}
\EndFor
\If{\tt{this} \bf{is} \tt{root}}
\State \tt{new.type= \sl{InternalBlock}-->\sl{RootBlock}}
\State \tt{new.size= max(root.blocks[i-1].\size { }+ new.num\sub{enq}- new.num\sub{deq}, 0)}\label{computeLength}
\EndIf

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i$\geq 1$}}
\Function{element}{GetEnqueue}{\sl{int} b, \sl{int} i} \cmt{Returns the \tt{element} of $E_\tt{i}(\tt{this},\tt{b})$.}
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element} \label{getBaseCase}
\ElsIf{\tt{i <= blocks[b].num\sub{enq-left}}} \label{leftOrRight} \cmt{$E_\tt{i}(\tt{this},\tt{b})$ is in the left child of this node.}
\State \tt{subblockIndex= left.BinarySearch(i+blocks[b-1].sum\sub{enq-left}, blocks[b-1].\eleft+1,} \label{leftChildGet}
\Statex \hspace{10.7em}\tt{blocks[b].\eleft)} 
\State \Return\tt{left.}\Call{GetEnqueue}{subblockIndex, i} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subblockIndex= right.BinarySearch(i+blocks[b-1].sum\sub{enq-right}, blocks[b-1].\eright+1,} \label{rightChildGet}
\Statex \hspace{11.1em}\tt{blocks[b].\eright)} 
\State \Return\tt{right.}\Call{GetEnqueue}{subblockIndex, i} 
\EndIf
\EndFunction{GetEnqueue}
